---
title: "HW2"
output: html_document
date: "2024-01-31"
---

```{r}
## load prostate data
prostate <- read.csv("/Users/margokim/Downloads/prostate.csv")

## subset to training examples
prostate_train <- subset(prostate, train==TRUE)

## plot lcavol vs lpsa
plot_psa_data <- function(dat=prostate_train) {
  plot(dat$lpsa, dat$lcavol,
       xlab="log Prostate Screening Antigen (psa)",
       ylab="log Cancer Volume (lcavol)",
       pch = 20)
  
}
plot_psa_data() 
```


1. Write functions that implement the L1 loss and tilted absolute loss functions.

```{r}
L1_loss <- function(y, yhat){
  return(abs(y - yhat))
}

## Tilted absolute loss functions
tilted_abs_loss_2 <- function(y, yhat, tau){
  if (y > yhat){
    return(tau * abs(y - yhat))
  } else {
    return((tau - 1) * abs(y - yhat))
  }
}
```


2. Create a figure that shows lpsa (x-axis) versus lcavol (y-axis). Add and label (e.g., using the 'legend' function in R) the linear model predictors associated with L2 loss, L1 loss, and tilted absolute value loss for tau = 0.25 and 0.75.

```{r}

## L2 loss function
L2_loss <- function(y, yhat){
  (y-yhat)^2 
}

## fit simple linear model using numerical optimization
## ... - arguments passed to los
fit_lin <- function(y, x, loss=L2_loss, beta_init = c(-0.51, 0.75), ...) {
  
  ## function to compute training error
  err <- function(beta)
    mean(loss(y,  beta[1] + beta[2]*x, ...))
  
  ## find value of beta that minimizes training error
  beta <- optim(par = beta_init, fn = err)
  
  
  return(beta)
}

## make predictions from linear model
predict_lin <- function(x, beta)
  beta[1] + beta[2]*x

## fit linear model
lin_beta <- fit_lin(y=prostate_train$lcavol,
                    x=prostate_train$lpsa,
                    loss=L2_loss)

## compute predictions for a grid of inputs
x_grid <- seq(min(prostate_train$lpsa),
              max(prostate_train$lpsa),
              length.out=100)
lin_pred <- predict_lin(x=x_grid, beta=lin_beta$par)

## plot data
plot_psa_data()

## plot predictions
lines(x=x_grid, y=lin_pred, col='darkgreen', lwd=2)

## do the same thing with 'lm'
lin_fit_lm <- lm(lcavol ~ lpsa, data=prostate_train)

## make predictins using 'lm' object
lin_pred_lm <- predict(lin_fit_lm, data.frame(lpsa=x_grid))

## plot predictions from 'lm'
lines(x=x_grid, y=lin_pred_lm, col='pink', lty=2, lwd=2)


##################################
## try modifying the loss function
##################################
## L1 loss function
L1_loss <- function(y, yhat){
  return(abs(y - yhat))
}

## Tilted absolute loss functions
tilted_abs_loss_2 <- function(y, yhat, tau){
  if (y > yhat){
    return(tau * abs(y - yhat))
  } else {
    return((tau - 1) * abs(y - yhat))
  }
}

## tilted absolute loss
tilted_abs_loss <- function(y, yhat, tau) {
  
  d <- y-yhat
  
  ifelse(d > 0, d * tau, d * (tau - 1))
}

custom_loss <- tilted_abs_loss

## plot custom loss function
err_grd <- seq(-1,1,length.out=200)
plot(err_grd, custom_loss(0, err_grd, tau=0.50), type='l',
     xlab='y-yhat', ylab='custom loss')

## fit linear model with custom loss
lin_beta_custom <- fit_lin(y=prostate_train$lcavol,
                           x=prostate_train$lpsa,
                           loss=custom_loss,
                           tau=0.25)

lin_pred_custom <- predict_lin(x=x_grid, beta=lin_beta_custom$par)

# Fit linear model with L1 loss
lin_beta_l1 <- fit_lin(y=prostate_train$lcavol, x=prostate_train$lpsa, loss=L1_loss)
lin_pred_l1 <- predict_lin(x=x_grid, beta=lin_beta_l1$par)

# Fit linear models with tilted absolute loss for tau = 0.75
lin_beta_tilted_25 <- fit_lin(y=prostate_train$lcavol, x=prostate_train$lpsa, loss=tilted_abs_loss, tau=0.25)
lin_pred_tilted_25 <- predict_lin(x=x_grid, beta=lin_beta_tilted_25$par)

lin_beta_tilted_75 <- fit_lin(y=prostate_train$lcavol, x=prostate_train$lpsa, loss=tilted_abs_loss, tau=0.75)
lin_pred_tilted_75 <- predict_lin(x=x_grid, beta=lin_beta_tilted_75$par)

## plot data
plot_psa_data()

## plot predictions from L2 loss
lines(x=x_grid, y=lin_pred, col='darkgreen', lwd=2)

## plot predictions from L1 loss
lines(x=x_grid, y=lin_pred_l1, col='red', lwd=2)

## plot predictions from tilted loss for tau = 0.25
lines(x=x_grid, y=lin_pred_tilted_25, col='pink', lwd=2, lty=2)

## plot predictions from tilted loss for tau = 0.75
lines(x=x_grid, y=lin_pred_tilted_75, col='orange', lwd=2, lty=3)

legend("topright", 
       legend=c("L2 Loss", "L1 Loss", "Tilted Loss (tau=0.25)", "Tilted Loss (tau=0.75)"),
       col=c("darkgreen", "red", "pink", "orange"), 
       lty=c(1, 1, 2, 3), 
       lwd=2)


```

3. Write functions to fit and predict from a simple nonlinear model with three parameters defined by 'beta[1] + beta[2]*exp(-beta[3]*x)'. Hint: make copies of 'fit_lin' and 'predict_lin' and modify them to fit the nonlinear model. Use c(-1.0, 0.0, -0.3) as 'beta_init'.


```{r}
fit_non_lin <- function(y, x, loss=L2_loss, beta_init = c(-1.0, 0.0, -0.3), ...) {
  
  ## function to compute training error
  err <- function(beta)
    mean(loss(y,  beta[1] + beta[2]*exp(-beta[3]*x), ...))
  
  ## find value of beta that minimizes training error
  beta <- optim(par = beta_init, fn = err)
  
  
  return(beta)
}

## make predictions from linear model
predict_non_lin <- function(x, beta) {
  beta[1] + beta[2]*exp(-beta[3]*x)
}

  
```

4. Create a figure that shows lpsa (x-axis) versus lcavol (y-axis). Add and label the nonlinear model predictions associated with L2 loss, L1 loss, and tilted absolute value loss for tau = 0.25 and 0.75.

```{r}
# Fit nonlinear model with L2 loss
non_lin_beta_L2 <- fit_non_lin(y=prostate_train$lcavol, x=prostate_train$lpsa, loss=L2_loss)

# Fit nonlinear model with L1 loss
non_lin_beta_L1 <- fit_non_lin(y=prostate_train$lcavol, x=prostate_train$lpsa, loss=L1_loss)

# Fit nonlinear models with tilted absolute loss for tau = 0.25 and 0.75
non_lin_beta_tilted_25 <- fit_non_lin(y=prostate_train$lcavol, x=prostate_train$lpsa, loss=tilted_abs_loss, tau=0.25)
non_lin_beta_tilted_75 <- fit_non_lin(y=prostate_train$lcavol, x=prostate_train$lpsa, loss=tilted_abs_loss, tau=0.75)


# Create a grid of x values for plotting predictions
x_grid <- seq(min(prostate_train$lpsa), max(prostate_train$lpsa), length.out=100)

# Plot the data
plot(prostate_train$lpsa, prostate_train$lcavol, 
     xlab="lpsa", ylab="lcavol", pch = 20, col = 'black')

# Add nonlinear predictions for each loss function
lines(x_grid, predict_non_lin(x_grid, non_lin_beta_L2$par), col='blue', lwd=2)
lines(x_grid, predict_non_lin(x_grid, non_lin_beta_L1$par), col='red', lwd=2)
lines(x_grid, predict_non_lin(x_grid, non_lin_beta_tilted_25$par), col='green', lwd=2, lty=2)
lines(x_grid, predict_non_lin(x_grid, non_lin_beta_tilted_75$par), col='orange', lwd=2, lty=3)

# Add a legend
legend("topright", 
       legend=c("Nonlinear L2 Loss", "Nonlinear L1 Loss", "Nonlinear Tilted Loss (tau=0.25)", "Nonlinear Tilted Loss (tau=0.75)"),
       col=c("blue", "red", "green", "orange"), 
       lty=c(1, 1, 2, 3), 
       lwd=2)




  
```
